The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-06 paper/iphas-dr2.tex
Update iphas-dr2.tex
Could you please add the following acknowledgment for me and Boris?

The research leading to these results has received funding from the
European Research Council under the European Union's Seventh Framework
Programme (FP/2007-2013) / ERC Grant Agreement n. 320964 (WDTracer).
BTG was supported in part by the UK’s Science and Technology
Facilities Council (ST/I001719/1)."

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 deg$^2$ imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against SDSS,
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure 
of our own backyard, the Milky Way.
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma -- 
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999}, 
to be extended into larger, deeper, more statistically-robust samples.
This in turn could better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we briefly remind of some key points.
IPHAS is a 1800~sq. deg. imaging survey of the northern Galactic Plane, 
providing images and photometry in Sloan $r, i$ 
along with narrowband H$\alpha$. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma. 
It is the first digital survey to offer comprehensive CCD photometry
of point sources in the Galactic Plane at visible wavelengths, 
and does so down to a limiting magnitude of $\sim$20th.
The IPHAS footprint on the northern sky spans a box 
of roughly 180 by 10 degrees, 
taking in the entire northern Galactic Plane 
at latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

The Wide Field Camera is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29 square degrees.
To cover the entire Northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
at a displacement of $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
The basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
The survey hence contains 15\,270 fields grouped into 7\,635 field pairs.
To achieve the desired survey depth of 20th magnitude or fainter, 
the filter exposure times were set at 120 sec (narrowband \ha), 
30 sec ($r$) and 10 sec ($i$)
in the majority of the survey observations.\footnote{In 2003 
the $r$-band exposure time was 10~sec instead of 30~sec,
and since Oct 2010 the $i$-band exposure time 
has been increased from 10~sec to 20~sec.}

@@

Figure~\ref{fig:footprint} shows the footprint 
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height (i.e. a square 0.33$\times$0.33\arcsec\ aperture)
as well as a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied which attempts to remove
the contamination from any nearby sources.
Whilst this approach works very well 
across the vast majority of the survey area,
the Galactic Plane unavoidably contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
in which case aperture photometry must always be interpreted 
with caution.
In \S\ref{sec:catalogue} we will explain that objects
which suffer from this problem 
are flagged in the catalogue using the \emph{deblend} warning flag.

Finally, an astrometric solution is determined
based on the 2MASS point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from a set of standard stars observed in the same night.
Sources are classified morphologically
-- stellar, galaxy or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

@@

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Figure~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\%of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

@@

The zeropoint was determined such that the resulting magnitude system
is in the WFC system that uses the SED of Vega 
as the zero colour reference object. 
Colour equations were used to transform between the IPHAS passbands 
and the Johnson-Cousins system 
of the published standard star photometry.
The entire procedure has been found to deliver zeropoints which 
are accurate at the level of 1-2\% 
in photometric conditions \citep{Gonzalez-Solares2011}.

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
at the level of 0.1 mag
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
Such variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of the Two Micron All Sky Survey \citep[2MASS;][]{Nikolaev2000},
the Sloan Digital Sky Survey \citep[SDSS;][]{Padmanabhan2008}
and the Panoramic Survey Telescope 
And Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

@@

\begin{table}
        \begin{center}
                \begin{tabular}{lcc}
                        \toprule
                         {\bf Before re-calibration} & Mean & $\sigma$  \\
                        \midrule
                        $r$ (IPHAS - APASS) & +0.014 & 0.104 \\
                        $i$ (IPHAS - APASS) & +0.007 & 0.108 \\
                        $r$ (IPHAS - SDSS) & +0.016 & 0.088 \\
                        $i$ (IPHAS - SDSS) & +0.010 & 0.089 \\
                        \bottomrule
                \end{tabular}
        \caption{Mean magnitude offsets for objects
                 cross-matched between IPHAS and APASS/SDSS
                 before the uniform re-calibration.
                 Eqns.~\ref{eqn:apass_r}-\ref{eqn:sdss_i} were applied
                 to the APASS/SDSS magnitudes to bring them into the
                 Vega-based IPHAS system prior to computing
                 the offsets.)
                 }
        \label{tbl:offsets_before}

                \begin{tabular}{lcc}
                        \toprule
                         {\bf After re-calibration} & Mean & $\sigma$ \\
                        \midrule
                        $r$ (IPHAS - APASS) & +0.000 & 0.011\\
                        $i$ (IPHAS - APASS) & +0.000 & 0.011 \\
                        $r$ (IPHAS - SDSS)  & -0.001 & 0.029\\
                        $i$ (IPHAS - SDSS) & -0.002 & 0.032 \\
                        \bottomrule
                \end{tabular}
        \caption{Same as Table~\ref{tbl:offsets_before}
                 but computed after the uniform re-calibration
                 was carried out.}
        \label{tbl:offsets_after}
        \end{center}
\end{table}

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are pre-dominantly located 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme, explained below,
which minimised the magnitude offsets between stars
located in the overlap regions with neighbouring fields.
Although we explained earlier that these overlap regions
are prone to systematics, the use of APASS enabled us to
keep the zeropoints of roughly half of the fields fixed,
which avoids these systematics from combining to introduce
artificial gradients across the survey.
In the next section we will show this to be true by validating
our calibration against SDSS.

@@

Anyone interested in the alternative detections of a source
-- hereafter called the \emph{secondary} detections --
can nevertheless obtain this information in two ways.
To begin with, whenever a secondary detection was observed 
within 10 minutes of the primary,
we have included the identifier and the default photometry
of that secondary detection
in the catalogue for convenience
(columns \emph{sourceID2}, \emph{fieldID2}, 
\emph{r2}, \emph{i2}, \emph{ha2},
\emph{rErr2}, \emph{iErr2}, \emph{haErr2}, \emph{errBits2}).
Such a secondary detection is available for 66\% of the sources brighter than $r<20$
due to the IPHAS field pair strategy.
In addition, the complete set of detection tables -- one for each exposure -- 
is made accessibly on our website to allow other uses of the data.
A user-friendly catalogue of secondary detections 
has not been compiled at present 
but may be part of a future data release.

@@

\begin{table*}
\vspace{2cm}

\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
  rErr\,$<$\,0.1 {\sc and} iErr\,$<$\,0.1 {\sc and} haErr\,$<$\,0.1 &
  109 million (50\%) &
  Require the photon noise to be less than 
  0.1 mag in all bands (i.e. S/N$>$10).
  This implicitly requires a detection in all three bands.  \\
  $r>13$ {\sc and} $i>12$ {\sc and} \ha\,$>12.5$ {\sc and not} \emph{saturated} &
  158 million (72\%) &
  The brightness must not exceed the nominal saturation limit
  and the peak pixel height must not exceed 55\,000 counts.
  Again, this implicitly requires a detection in all three bands.
  \\
  $|\mathrm{r}- \mathrm{rAperMag1}| 
  < 3\sqrt{\mathrm{rErr}^2 + \mathrm{rAperMag1Err}^2} 
  + 0.03$ &
  176 million (80\%) &
  Require the $r$ magnitude measured 
  in the default 2.3\arcsec\ diameter aperture
  to be consistent with the measurement 
  made in the smaller 1.2\arcsec\ aperture,
  albeit tolerating a 0.03 mag systematic error.
  This will reject sources for which the background
  subtraction or the deblending procedure
  was not performed reliably. \\
  $|\mathrm{i}- \mathrm{iAperMag1}| 
  < 3\sqrt{\mathrm{iErr}^2 + \mathrm{iAperMag1Err}^2} 
  + 0.03$ &
  183 million (84\%) &
  Same as above for $i$. \\
  $|\mathrm{ha}- \mathrm{haAperMag1}| < 
  3\sqrt{\mathrm{haErr}^2 + \mathrm{haAperMag1Err}^2} 
  + 0.03$ &
  158 million (72\%) &
  Same as above for \ha. \\
  \hline
  All of the above (flagged as {\bf\emph{reliable}}) &
  86 million (39\%) & \\
  \hline
\end{tabular}
\caption{Recommended minimum quality criteria 
for selecting objects with reliable colours 
from the IPHAS DR2 source catalogue. 
86~million entries in the catalogue (39\%)
satisfy all the criteria listed in this table.
For convenience, these have been flagged in the catalogue
using the column named \emph{reliable}.}
\label{tab:reliable}

\vspace{2cm}

\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
   reliable &
   86 million (39\%) &
   The object must satisfy the criteria listed in Table~\ref{tab:reliable}. \\

@@

  All of the above (flagged as {\bf\emph{veryReliable}}) &
  59 million (27\%) & \\
  \hline
\end{tabular}
\caption{Additional quality criteria which are recommended
for applications which require very reliable colours
at the expense of completeness. 
For convenience, the sources which satisfy the criteria listed
in this table have been flagged in the catalogue
using the column named \emph{veryReliable}.}
\label{tab:veryreliable}

@@

In addition to the global trend, 
there are significant variations
in the source density on smaller scales.
For example,  we find a significant drop near the constellations 
of Cygnus ($l\simeq 80$) and Aquila ($l\simeq 40$),
which are regions known to be affected
by high levels of foreground extinction.
Dark clouds are readily visible towards these constellation by eye,
and they have often been referred to as ``the Great Rift''.

@@

The survey's unique $(r-\ha)$ colour,
when combined with $(r-i)$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
Put differently, the main sequence in the $(r-\ha,\ r-i)$ diagram
runs in a direction that is different from the reddening vector,
because the $r$-\ha\ colour tends to act
as a coarse proxy for spectral type
and is less sensitive to reddening than $(r-i)$.
As a result, the distribution of a stellar population
in the IPHAS colour-colour diagram
can offer a handle on the properties of the population
and the extinction along a line of sight.

>>>>

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 deg$^2$ imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against SDSS,
% RR: better Sloan Digital Sky Survey (SDSS)?
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure 
of our own backyard, the Milky Way.
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma -- 
% RR: The INT acronym is defined before
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999} 
to be extended into larger, deeper, and more statistically-robust samples.
This in turn could better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release, we briefly remind of some key points.
IPHAS is a 1800~deg$^{2}$. imaging survey of the northern Galactic Plane, 
providing images and photometry in Sloan $r, i$ 
along with narrowband H$\alpha$. 
It is carried out using the WFC 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive CCD photometry
of point sources in the Galactic Plane at visible wavelengths, 
and does so down to a limiting magnitude of $\sim$20th.
The IPHAS footprint spans a box 
of roughly 180 by 10 degrees, 
taking in the entire northern Galactic Plane 
at latitudes $-5^{\circ} < b < +5^{\circ}$ 
and longitudes $30^{\circ} < l < 215^{\circ}$.

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^{2}$.
To cover the entire Northern Plane with some overlap,
the survey area was divided into 7635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in declination 
and $+$5 arcmin in right ascension,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
The basic unit of observation hence
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
The survey hence contains 15\,270 fields grouped into 7\,635 field pairs.
To achieve the desired survey depth of 20th magnitude or fainter, 
the filter exposure times were set at 120 sec (narrowband \ha), 
30 sec ($r$) and 10 sec ($i$)
in the majority of the survey observations.\footnote{In 2003 
the $r$-band exposure time was 10~sec instead of 30~sec,
and since Oct 2010 the $i$-band exposure time 
has been increased from 10~sec to 20~sec.}

@@

Fig.~\ref{fig:footprint} shows the footprint % Should replace all Figure with Fig., according to the MNRAS style I think
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8\% of the survey area --
are predominantly located towards the Galactic anti-centre 
at $l > 120^o$.
Fields at these longitudes can only be accessed from La Palma 
in the months of November-December,
which is when the weather and seeing conditions are often poor
at the INT and observing attempts have failed repeatedly.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

The reduced images are then stored in multi-extension FITS files
with a primary header describing the characteristics
(position, filter, exposure time, etc.) 
and four image extensions 
corresponding to each of the four CCDs.
Source detection and characterisation is then carried out 
using the \textsc{imcore} tool \citep{Irwin1985,Irwin1997}.
The flux of each source is measured using both
the peak pixel height (i.e. a square 0.33$\arcsec\times$0.33\arcsec\ aperture)
% not sure if you need the \arcsec I added, but it would look nicer to me.
as well as a series of circular apertures of increasing diameter 
(1.2\arcsec, 2.3\arcsec, 3.3\arcsec, 4.6\arcsec\ and 6.6\arcsec).

The local background levels are estimated 
by computing the sigma-clipped median
flux in a grid of 64$\times$64 pixels (21\arcsec$\times$21\arcsec)
across the image,
which is then interpolated to obtain an estimate 
of the background level at each pixel.
These sky levels are subtracted from the aperture photometry and
-- when required --
a deblending routine is applied which attempts to remove
the contamination from any nearby sources.
Whilst this approach works very well 
across the vast majority of the survey area,
the Galactic Plane unavoidably contains crowded regions 
with large numbers of overlapping sources
or rapidly spatially-varying nebulosity,
in which case aperture photometry must always be interpreted 
with caution.
In \S\ref{sec:catalogue} we will explain that objects
which suffer from this problem 
are flagged in the catalogue using the \emph{deblend} warning flag.

Finally, an astrometric solution is determined
based on the Two-Micron All Sky Survey (2MASS) point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also provided 
based on the average zeropoint
determined from a set of standard stars observed in the same night.
Sources are classified morphologically
-- stellar, galaxy or noise --
based on the curve-of-growth determined
from measuring the source intensity in a series of growing apertures.
Finally, the resulting source detection tables are also stored 
in multi-extension FITS files.

@@

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Figure~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\% of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

@@

The zeropoint was determined such that the resulting magnitude system
is in the WFC system that uses the Spectral Energy Distribution (SED) of Vega 
as the zero colour reference object. 
Colour equations were used to transform between the IPHAS passbands 
and the Johnson-Cousins system 
of the published standard star photometry.
The entire procedure has been found to deliver zeropoints which 
are accurate at the level of 1-2\% 
in photometric conditions \citep{Gonzalez-Solares2011}.

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
at the level of 0.1 mag
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
Such variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of the Two Micron All Sky Survey \citep[][]{Nikolaev2000},
the Sloan Digital Sky Survey \citep[][]{Padmanabhan2008}
% Define both 2MASS and SDSS before in the text
and the Panoramic Survey Telescope 
And Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

@@

\begin{table}
% The caption of tables should be on top and not at the bottom
        \caption{Mean magnitude offsets for objects
                 cross-matched between IPHAS and APASS/SDSS
                 before the uniform re-calibration.
                 Eqns.~\ref{eqn:apass_r}-\ref{eqn:sdss_i} were applied
                 to the APASS/SDSS magnitudes to bring them into the
                 Vega-based IPHAS system prior to computing
                 the offsets.)
                 }
        \label{tbl:offsets_before}
        \begin{center}
                \begin{tabular}{lcc}
                        \toprule
                         {\bf Before re-calibration} & Mean & $\sigma$  \\
                        \midrule
                        $r$ (IPHAS - APASS) & +0.014 & 0.104 \\
                        $i$ (IPHAS - APASS) & +0.007 & 0.108 \\
                        $r$ (IPHAS - SDSS) & +0.016 & 0.088 \\
                        $i$ (IPHAS - SDSS) & +0.010 & 0.089 \\
                        \bottomrule
                \end{tabular}
        \caption{Same as Table~\ref{tbl:offsets_before}
                 but computed after the uniform re-calibration
                 was carried out.}
        \label{tbl:offsets_after}
                \begin{tabular}{lcc}
                        \toprule
                         {\bf After re-calibration} & Mean & $\sigma$ \\
                        \midrule
                        $r$ (IPHAS - APASS) & +0.000 & 0.011\\
                        $i$ (IPHAS - APASS) & +0.000 & 0.011 \\
                        $r$ (IPHAS - SDSS)  & -0.001 & 0.029\\
                        $i$ (IPHAS - SDSS) & -0.002 & 0.032 \\
                        \bottomrule
                \end{tabular}

        \end{center}
\end{table}

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are predominantly located 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme, explained below,
which minimised the magnitude offsets between stars
located in the overlap regions with neighbouring fields.
Although we explained earlier that these overlap regions
are prone to systematics, the use of APASS enabled us to
keep the zeropoints of roughly half of the fields fixed,
which avoids these systematics from combining to introduce
artificial gradients across the survey.
In the next section we will show this to be true by validating
our calibration against SDSS.

@@

Anyone interested in the alternative detections of a source
-- hereafter called the \emph{secondary} detections --
can nevertheless obtain this information in two ways.
To begin with, whenever a secondary detection was observed 
within 10 minutes of the primary,
we have included the identifier and the default photometry
of that secondary detection
in the catalogue for convenience
(columns \emph{sourceID2}, \emph{fieldID2}, 
\emph{r2}, \emph{i2}, \emph{ha2},
\emph{rErr2}, \emph{iErr2}, \emph{haErr2}, \emph{errBits2}).
Such a secondary detection is available for 66\% of the sources brighter than $r<20$
due to the IPHAS field pair strategy.
In addition, the complete set of detection tables -- one for each exposure -- 
is made accessible on our website to allow other uses of the data.
A user-friendly catalogue of secondary detections 
has not been compiled at present 
but may be part of a future data release.

@@

\begin{table*}

\vspace{2cm}
\caption{Recommended minimum quality criteria 
for selecting objects with reliable colours 
from the IPHAS DR2 source catalogue. 
86~million entries in the catalogue (39\%)
satisfy all the criteria listed in this table.
For convenience, these have been flagged in the catalogue
using the column named \emph{reliable}.
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
  rErr\,$<$\,0.1 {\sc and} iErr\,$<$\,0.1 {\sc and} haErr\,$<$\,0.1 &
  109 million (50\%) &
  Require the photon noise to be less than 
  0.1 mag in all bands (i.e. S/N$>$10).
  This implicitly requires a detection in all three bands.  \\
  $r>13$ {\sc and} $i>12$ {\sc and} \ha\,$>12.5$ {\sc and not} \emph{saturated} &
  158 million (72\%) &
  The brightness must not exceed the nominal saturation limit
  and the peak pixel height must not exceed 55\,000 counts.
  Again, this implicitly requires a detection in all three bands.
  \\
  $|\mathrm{r}- \mathrm{rAperMag1}| 
  < 3\sqrt{\mathrm{rErr}^2 + \mathrm{rAperMag1Err}^2} 
  + 0.03$ &
  176 million (80\%) &
  Require the $r$ magnitude measured 
  in the default 2.3\arcsec\ diameter aperture
  to be consistent with the measurement 
  made in the smaller 1.2\arcsec\ aperture,
  albeit tolerating a 0.03 mag systematic error.
  This will reject sources for which the background
  subtraction or the deblending procedure
  was not performed reliably. \\
  $|\mathrm{i}- \mathrm{iAperMag1}| 
  < 3\sqrt{\mathrm{iErr}^2 + \mathrm{iAperMag1Err}^2} 
  + 0.03$ &
  183 million (84\%) &
  Same as above for $i$. \\
  $|\mathrm{ha}- \mathrm{haAperMag1}| < 
  3\sqrt{\mathrm{haErr}^2 + \mathrm{haAperMag1Err}^2} 
  + 0.03$ &
  158 million (72\%) &
  Same as above for \ha. \\
  \hline
  All of the above (flagged as {\bf\emph{reliable}}) &
  86 million (39\%) & \\
  \hline
\end{tabular}
}
\label{tab:reliable}

\vspace{2cm}
\caption{Additional quality criteria which are recommended
for applications which require very reliable colours
at the expense of completeness. 
For convenience, the sources which satisfy the criteria listed
in this table have been flagged in the catalogue
using the column named \emph{veryReliable}.
\begin{tabular}{p{8cm}lp{6cm}}
  \hline
  Quality criterion & Rows passed & Description \\
  \hline
   reliable &
   86 million (39\%) &
   The object must satisfy the criteria listed in Table~\ref{tab:reliable}. \\

@@

  All of the above (flagged as {\bf\emph{veryReliable}}) &
  59 million (27\%) & \\
  \hline
\end{tabular}
}
\label{tab:veryreliable}

@@

In addition to the global trend, 
there are significant variations
in the source density on smaller scales.
For example,  we find a significant drop near the constellations 
of Cygnus ($l\simeq 80$) and Aquila ($l\simeq 40$),
% what's the low surface density seen at ~90 deg?
which are regions known to be affected
by high levels of foreground extinction.
Dark clouds are readily visible towards these constellation by eye,
and they have often been referred to as ``the Great Rift''.

@@

The survey's unique $(r-\ha)$ colour,
when combined with $(r-i)$,
has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
Put differently, the main sequence in the $(r-\ha,\ r-i)$ diagram
runs in a direction that is different from the reddening vector,
because the ($r-\ha$) colour tends to act
as a coarse proxy for spectral type
and is less sensitive to reddening than $(r-i)$.
As a result, the distribution of a stellar population
in the IPHAS colour-colour diagram
can offer a handle on the properties of the population
and the extinction along a line of sight.
